---
title: "Running q2e from R"
author: "Simon Hickinbotham"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Vignette Title}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```
```{r set-options, echo=FALSE, cache=FALSE}
options(width = 1200)
```


# Basic usage: 

A simple way of using q2e in your project is to run: 


```{r eval=FALSE, include=TRUE}
library(q2e)
runq2e("Datafile.txt")
```


This will:

- run the q2e analysis on "Datafile.txt" - and 
- create two output files in the current working directory
    - `rq2eResults.txt`
    - `r2e2Betas.csv`
  
The input file must have two columns of numbers representing the mass and intensity of a MALDI spectrum. Data files must separate these values with whitespace, but files ending `.csv` must be comma-separated. 

The output files are identical to the files produced in the original C++ code. 


# Detailed usage



This



# 

The R version of q2e runs in *almost* the same way as Julie Wilson's original C version. The basic idea was that the list of MALDI data files to be processed, the peptides to check against and the the parameters of the algorithm are contained in a number of files. In the R version, the datafiles are passed in as a vector of file paths, and default values for the peptides and parameters are provided.

To illustrate this, let's load the datafiles provided with the q2e package and run some analyses. 

## Load the example data files. 


The datafiles are space delimited and consist of two columns: the mass and the intensity of the MALDI spectrum, like this: 




```{r}
  datafiles <- c(system.file("extdata", "1052a1_65_1_plastic_M19_04132012Kinetics1rr.txt", package = "q2e"),
                 system.file("extdata", "1052a2_65_1_plastic_A1_04132012Kinetics1rr.txt", package = "q2e"),
                 system.file("extdata", "1052a3_65_1_plastic_A16_04132012Kinetics1rr.txt", package = "q2e"),
                 system.file("extdata", "1053b1_65_1_plastic_M20_04132012Kinetics1rr.txt", package = "q2e"),
                 system.file("extdata", "1053b2_65_1_plastic_A2_04132012Kinetics1rr.txt", package = "q2e"),
                 system.file("extdata", "1053b3_65_1_plastic_A17_04132012Kinetics1rr.txt", package = "q2e"))
```




## Running the basic program. 


# Vignettes

Vignettes are long form documentation commonly included in packages. Because they are part of the distribution of the package, they need to be as compact as possible. The `html_vignette` output type provides a custom style sheet (and tweaks some options) to ensure that the resulting html is as small as possible. The `html_vignette` format:

- Never uses retina figures
- Has a smaller default figure size
- Uses a custom CSS stylesheet instead of the default Twitter Bootstrap style

## Vignette Info

Note the various macros within the `vignette` section of the metadata block above. These are required in order to instruct R how to build the vignette. Note that you should change the `title` field and the `\VignetteIndexEntry` to match the title of your vignette.

## Styles

The `html_vignette` template includes a basic CSS theme. To override this theme you can specify your own CSS in the document metadata as follows:

    output: 
      rmarkdown::html_vignette:
        css: mystyles.css

## Figures

The figure sizes have been customised so that you can easily put two images side-by-side. 

```{r, fig.show='hold'}
plot(1:10)
plot(10:1)
```

You can enable figure captions by `fig_caption: yes` in YAML:

    output:
      rmarkdown::html_vignette:
        fig_caption: yes

Then you can use the chunk option `fig.cap = "Your figure caption."` in **knitr**.

## More Examples

You can write math expressions, e.g. $Y = X\beta + \epsilon$, footnotes^[A footnote here.], and tables, e.g. using `knitr::kable()`.

```{r, echo=FALSE, results='asis'}
knitr::kable(head(mtcars, 10))
```

Also a quote using `>`:

> "He who gives up [code] safety for [code] speed deserves neither."
([via](https://twitter.com/hadleywickham/status/504368538874703872))
